The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-09-30 paper.tex
All edits to Irish -- DONE

\ex. Ident(SonAngle)$<$1.5 $\gg$ {\sc *ComplexCoda} $\gg$ {\sc Dep}
     \a. Allows epenthesis in rising and level clusters such as stop-fricative TF:
\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/TF/} & {\sc Ident(Son$\measuredangle$)}$<$1.5 & {\sc *ComplexCoda} & {\sc Dep} \\[0.5ex]
\hline \hline a. & & TF & & $\ast$! & \cellcolor{lightgray} \\
\hline b. & \ding{43} & T\textipa{@}F & 0.59 & & \cellcolor{lightgray}$\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}
     \b. Prevents epenthesis in falling clusters such as FT:
\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/FT/} & {\sc Ident(Son$\measuredangle$)}$<$1.5 & {\sc *ComplexCoda} & {\sc Dep} \\[0.5ex]
\hline \hline c. & \ding{43} & FT & &\cellcolor{lightgray} $\ast$ & \cellcolor{lightgray} \\
\hline d. & & F\textipa{@}T & 2.11 $\ast$! &\cellcolor{lightgray} & \cellcolor{lightgray}$\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

@@

@@

\ex. {\sc Sonority Angles} of sonorant-initial clusters using revised sonority scale

@@

        \node (NT) at (2.48619449019234 * 15, 0) {NT}; % 2.35619449019234 - repositioning for visual clarity
        \node (ND) at (2.28183949564558 * 15, 0) {ND}; % 2.23183949564558 
        \node (NF) at (2.03444393579570 * 15, 0) {NF};
        \node (NN) at (1.24904577239825 * 15, 0) {NN};
        \node (NR) at (0.60 * 15, 0) {NR}; % 0.4636476090008

        \node (RT) at (2.48619449019234 * 15, 0.7) {RT};% 2.35619449019234 - repositioning for visual clarity
        \node (RD) at (2.37743866747662 * 15, 0) {RD}; % 2.29743866747662 
        \node (RF) at (2.19429743558818 * 15, 0) {RF}; % 2.21429743558818 
        \node (RN) at (1.89254688119154 * 15, 0) {RN};
        \node (RR) at (1.10714871779409 * 15, 0) {RR};

@@

  \node (2pointlabel) at (2.43 * 15,-1.7) {2.3};

@@

\ex. {\sc Ident(SonAngle)$<$2.3} $\gg$ {\sc *Son-C} $\gg$ {\sc Ident(SonAngle)$<$2.4}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{kork}/} & {\sc Ident(SonAngle)$<$2.3} & {\sc *Son-C} & {\sc Ident(SonAngle)$<$2.4} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{kork} & & \cellcolor{lightgray}$\ast$ & \cellcolor{lightgray} \\
\hline b. & & \textipa{kor@k} & 2.35 $\ast$! & \cellcolor{lightgray} & \cellcolor{lightgray}2.35 $\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{alb@}/} & {\sc Ident(SonAngle)$<$2.3} & {\sc *Son-C} & {\sc Ident(SonAngle)$<$2.4} \\[0.5ex]
\hline \hline c. & \ding{43} & \textipa{al@b@} & 2.29 & & \cellcolor{lightgray} 2.29 \\
\hline d. & & \textipa{alb@} & & $\ast$! & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

@@

Using the same revised sonority scale, the {\sc Sonority Rise} calculations are as follows for the relevant clusters:

@@

        \node (NT) at (1.66666666666667 * 15, 0) {NT}; 
        \node (ND) at (1.5 * 15, 0) {ND}; 
        \node (NF) at (1.33333333333333 * 15, 0) {NF};
        \node (NN) at (1 * 15, 0) {NN};
        \node (NR) at (0.666666666666667 * 15, 0) {NR};

        \node (RT) at (2.5 * 15, 0) {RT};
        \node (RD) at (2.25 * 15, 0) {RD}; 
        \node (RF) at (2 * 15, 0) {RF}; 
        \node (RN) at (1.5 * 15, 0.7) {RN};
        \node (RR) at (1 * 15, 0.7) {RR};

    \node (axisstart) at (0.5 * 15,-1) {};
    \node (axisend)   at (2.6 * 15,-1) {};
    \draw (axisstart) -- (axisend);
   \node (xaxislabel) at (2.6 * 15,-1.5) {\textsc{Sonority Rise}};

@@

Note that the clusters RF and RD fall between RT and NT on this scale.  Therefore, we cannot clearly separate the clusters that undergo epenthesis
from the ones that do not if we adopt {\sc Sonority Rise} as the relevant metric.

@@

Carnie's (1994) analysis of these facts is in terms of a Minimal Distancing Constraint with various parameter settings, such as for which direction away from the nucleus it applies (to the right in the Irish case, to account for epenthesis applying to both codas and heterosyllabic clusters). His preliminary analysis has distance computed according to the following sonority scale:

\ex. \label{carniesonorityscale} \a. 1 - Voiceless stops
     \b. 2 - Voiced stops
     \b. 3 - Voiceless fricatives
     \b. 4 - Voiced fricatives
     \b. 5 - Sonorants

with a subsequent revision where sonority is computed in terms of the number of nodes that have to be specified to obtain the appropriate manner specification. This computation is somewhat complicated, so I will not reproduce it here. The result is that sonorant-fricative and sonorant-voiced stop clusters have a sonority distance of 0, since they are each specified for a single node in the feature geometry. Voiceless stops have no specified nodes, so sonorant-voiceless stop clusters have a sonority distance of 1, which is the minimal distance tolerated.

A sonority scale that gives nasals and liquids identical sonority values seems unlikely given measurable differences in the phonetic correlates of sonority such as intensity between liquids and nasals \citep{parker.2002}. This is especially so when the obstruents are spread so widely on the sonority scale. 
I will therefore explore instead an alternative sonority distance-based analysis using Syllable Contact.

We quickly see that Syllable Contact on its own is insufficient as epenthesis in Irish also affects coda clusters. We will need to invoke Sonority Sequencing.

@@

Suppose we define a single markedness constraint that favours all falling sonority clusters regardless of their position in the syllable, i.e. combining Syllable Contact and Sonority Sequencing. We require a finer-grained metric to measure how marked a particular cluster is. An obvious course of action is to use \citet{gouskova.2002}'s {\sc *Dis} metric, extending its use from the heterosyllabic environment alone to all positions.

@@

@@

        \node (NT) at (2 * 5, 0) {NT}; 
        \node (ND) at (1.5 * 5, 0) {ND}; 
        \node (NF) at (1 * 5, 0) {NF};
        \node (NN) at (0 * 5, 0) {NN};
        \node (NR) at (-1 * 5, 0) {NR};

        \node (RT) at (3 * 5, 0) {RT};
        \node (RD) at (2.5 * 5, 0) {RD}; 
        \node (RF) at (2 * 5, 0.7) {RF}; 
        \node (RN) at (1 * 5, 0.7) {RN};
        \node (RR) at (0 * 5, 0.7) {RR};

>>>>

\ex. Ident(Son$\measuredangle$)$<$1.5 $\gg$ {\sc *ComplexCoda} $\gg$ {\sc Dep}
     \a. Allows epenthesis in rising and level clusters such as stop-fricative TF:
\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/TF/} & {\sc Ident(Son$\measuredangle$)}$<$1.5 & {\sc *ComplexCoda} & {\sc Dep} \\[0.5ex]
\hline \hline a. & & TF & & $\ast$! & \cellcolor{lightgray} \\
\hline b. & \ding{43} & T\textipa{@}F & 0.59 & & \cellcolor{lightgray}$\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}
     \b. Prevents epenthesis in falling clusters such as FT:
\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/FT/} & {\sc Ident(Son$\measuredangle$)}$<$1.5 & {\sc *ComplexCoda} & {\sc Dep} \\[0.5ex]
\hline \hline c. & \ding{43} & FT & &\cellcolor{lightgray} $\ast$ & \cellcolor{lightgray} \\
\hline d. & & F\textipa{@}T & 2.11 $\ast$! &\cellcolor{lightgray} & \cellcolor{lightgray}$\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

@@

This constraint does not duplicate the work of the {\sc Ident(Son$\measuredangle$)}$<$n constraints, but eliminates the obstruent-initial clusters from consideration, as these are not targeted for epenthesis by any high-ranking constraints.

@@

\ex. {\sc Sonority Angles} of sonorant-initial clusters using revised sonority scale \label{sonorantinitial_sonangle}

@@

        \node (RT) at (2.48619449019234 * 15, 0.7) {RT};% 2.35619449019234 - repositioning for visual clarity
        \node (NT) at (2.48619449019234 * 15, 0) {NT}; % 2.35619449019234 - repositioning for visual clarity
        \node[font=\footnotesize,opacity=0.7] (NTno) at (2.48619449019234 * 15, 1.5) {2.36};
        \node (RD) at (2.37743866747662 * 15, 0) {RD}; % 2.29743866747662 
        \node[font=\footnotesize,opacity=0.7] (RDno) at (2.37743866747662 * 15, 1.5) {2.30};
        \node (ND) at (2.28183949564558 * 15, 0) {ND}; % 2.23183949564558 
        \node[font=\footnotesize,opacity=0.7] (NDno) at (2.28183949564558 * 15, 1.5) {2.23};
        \node (RF) at (2.19429743558818 * 15, 0) {RF}; % 2.21429743558818 
        \node[font=\footnotesize,opacity=0.7] (RFno) at (2.19429743558818 * 15, 1.5) {2.21};
        \node (NF) at (2.03444393579570 * 15, 0) {NF};
        \node[font=\footnotesize,opacity=0.7] (NFno) at (2.03444393579570 * 15, 1.5) {2.03};
        \node (RN) at (1.89254688119154 * 15, 0) {RN};
        \node[font=\footnotesize,opacity=0.7] (RNno) at (1.89254688119154 * 15, 1.5) {1.89};
        \node (NN) at (1.24904577239825 * 15, 0) {NN};
        \node[font=\footnotesize,opacity=0.7] (NNno) at (1.24904577239825 * 15, 1.5) {1.25};
        \node (RR) at (1.10714871779409 * 15, 0) {RR};
        \node[font=\footnotesize,opacity=0.7] (RRno) at (1.10714871779409 * 15, 1.5) {1.11};
        \node (NR) at (0.60 * 15, 0) {NR}; % 0.4636476090008
        \node[font=\footnotesize,opacity=0.7] (NRno) at (0.60 * 15, 1.5) {0.46};

@@

  \node (2pointlabel) at (2.43 * 15,-1.7) {2.33};

@@

\ex. {\sc Ident(Son$\measuredangle$)$<$2.33} $\gg$ {\sc *Son-C} $\gg$ {\sc Ident(Son$\measuredangle$)$<$2.4}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{kork}/} & {\sc Ident(Son$\measuredangle$)$<$2.33} & {\sc *Son-C} & {\sc Ident(Son$\measuredangle$)$<$2.4} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{kork} & & \cellcolor{lightgray}$\ast$ & \cellcolor{lightgray} \\
\hline b. & & \textipa{kor@k} & 2.36 $\ast$! & \cellcolor{lightgray} & \cellcolor{lightgray}2.36 $\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{alb@}/} & {\sc Ident(Son$\measuredangle$)$<$2.33} & {\sc *Son-C} & {\sc Ident(Son$\measuredangle$)$<$2.4} \\[0.5ex]
\hline \hline c. & \ding{43} & \textipa{al@b@} & 2.29 & & \cellcolor{lightgray} 2.29 \\
\hline d. & & \textipa{alb@} & & $\ast$! & \cellcolor{lightgray} \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

@@

An analysis based on {\sc Sonority Rise} \citep{flemming.2008} would be largely identical, but the cluster hierarchy would differ:

@@

        \node[draw,circle,dashed] (NT) at (1.66666666666667 * 15, 0) {NT}; 
        \node[font=\footnotesize,opacity=0.7] (NTno) at (1.66666666666667  * 15, 1.5) {1.67};
        \node (ND) at (1.5 * 15, 0) {ND}; 
        \node[font=\footnotesize,opacity=0.7] (NDno) at (1.5 * 15, 1.5) {1.5};
        \node (NF) at (1.33333333333333 * 15, 0) {NF};
        \node[font=\footnotesize,opacity=0.7] (NDno) at (1.33 * 15, 1.5) {1.33};
        \node (NN) at (1 * 15, 0) {NN};
        \node[font=\footnotesize,opacity=0.7] (NDno) at (1 * 15, 1.5) {1.0};
        \node (NR) at (0.666666666666667 * 15, 0) {NR}; 
        \node[font=\footnotesize,opacity=0.7] (NDno) at (0.666666666666667* 15, 1.5) {0.67};

        \node[draw,circle,dashed] (RT) at (2.5 * 15, 0) {RT};
        \node[font=\footnotesize,opacity=0.7] (RTno) at (2.5 * 15, 1.5) {2.5};
        \node (RD) at (2.25 * 15, 0) {RD}; 
        \node[font=\footnotesize,opacity=0.7] (RDno) at (2.25 * 15, 1.5) {2.25};
        \node (RF) at (2 * 15, 0) {RF}; 
        \node[font=\footnotesize,opacity=0.7] (RFno) at (2 * 15, 1.5) {2.0};
        \node (RN) at (1.5 * 15, 0.7) {RN};
        \node (RR) at (1 * 15, 0.7) {RR};

    \node (axisstart) at (0.5 * 15,-1.5) {};
    \node (axisend)   at (2.6 * 15,-1.5) {};
    \draw (axisstart) -- (axisend);
   \node (xaxislabel) at (2.6 * 15,-2.0) {\textsc{Sonority Rise}};

@@

Note that the clusters RF and RD fall between RT and NT on this scale.  Therefore, we cannot separate the clusters that undergo epenthesis from the ones that do not if we adopt {\sc Sonority Rise} as the relevant metric, as we can when we adopt {\sc Sonority Angle} (c.f. \ref{sonorantinitial_sonangle}).

@@

%Carnie's (1994) analysis of these facts is in terms of a Minimal Distancing Constraint with various parameter settings, such as for which direction away from the nucleus it applies (to the right in the Irish case, to account for epenthesis applying to both codas and heterosyllabic clusters). His preliminary analysis has distance computed according to the following sonority scale:
%
%\ex. \label{carniesonorityscale} \a. 1 - Voiceless stops
%     \b. 2 - Voiced stops
%     \b. 3 - Voiceless fricatives
%     \b. 4 - Voiced fricatives
%     \b. 5 - Sonorants
%
%with a subsequent revision where sonority is computed in terms of the number of nodes that have to be specified to obtain the appropriate manner specification. This computation is somewhat complicated, so I will not reproduce it here. The result is that sonorant-fricative and sonorant-voiced stop clusters have a sonority distance of 0, since they are each specified for a single node in the feature geometry. Voiceless stops have no specified nodes, so sonorant-voiceless stop clusters have a sonority distance of 1, which is the minimal distance tolerated. 
%
%A sonority scale that gives nasals and liquids identical sonority values seems unlikely given measurable differences in the phonetic correlates of sonority such as intensity between liquids and nasals \citep{parker.2002}. This is especially so when the obstruents are spread so widely on the sonority scale. 
%I will therefore explore instead an alternative sonority distance-based analysis using Syllable Contact.

Now let us construct a markedness-based analysis for the Irish data.\footnote{\cite{carnie.1994} analyses these facts with a Minimal Distancing Constraint, based on a feature-geometric notion of sonority distance. The analysis presented here adheres more closely to the phonetic interpretation of sonority (see, for example, \cite{parker.2002}.}

@@

Suppose then that we define a single markedness constraint that favours all falling sonority clusters regardless of their position in the syllable, i.e. combining Syllable Contact and Sonority Sequencing. We require a finer-grained metric to measure how marked a particular cluster is. An obvious course of action is to use \citet{gouskova.2002}'s {\sc Dis} metric, extending its use from the heterosyllabic environment alone to all positions.

@@

\ex. {\sc Dis} values of sonorant-initial clusters using revised sonority scale

@@

        \node[draw,circle,dashed] (NT) at (2 * 5, 0) {NT}; 
        \node[font=\footnotesize,opacity=0.7] (RTno) at (2 * 5, 2.5) {2};
        \node (ND) at (1.5 * 5, 0) {ND}; 
        \node[font=\footnotesize,opacity=0.7] (NDno) at (1.5 * 5, 2.5) {1.5};
        \node (NF) at (1 * 5, 0) {NF};
        \node[font=\footnotesize,opacity=0.7] (NFno) at (1 * 5, 2.5) {1};
        \node (NN) at (0 * 5, 0) {NN};
        \node[font=\footnotesize,opacity=0.7] (NNno) at (0 * 5, 2.5) {0};
        \node (NR) at (-1 * 5, 0) {NR}; 
        \node[font=\footnotesize,opacity=0.7] (NRno) at (-1 * 5, 2.5) {-1};

        \node[draw,circle,dashed] (RT) at (3 * 5, 0) {RT};
        \node[font=\footnotesize,opacity=0.7] (NDno) at (3 * 5, 2.5) {3};
        \node (RD) at (2.5 * 5, 0) {RD}; 
        \node[font=\footnotesize,opacity=0.7] (RDno) at (2.5 * 5, 2.5) {2.5};
        \node (RF) at (2 * 5, 1.5) {RF}; 
        \node (RN) at (1 * 5, 0.7) {RN};
        \node (RR) at (0 * 5, 0.7) {RR};
